Homicide Increases in the COVID-19 Era

### 2025-02-22 Draft1.qmd
added subtitle for bib

```{r, output=FALSE}
library(sf)
library(tidygeocoder)
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer)

@@

```{r}
# Calculate quantiles for the entire dataset (from 2018 to 2023)
quantiles <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

# Setting a layout for a 2x3 grid
par(mfrow = c(2, 3), mar = c(1, 1, 1, 5), oma = c(0, 0, 3, 0), cex.main = 1.5)

# Define color palette
palette <- brewer.pal(7, "Greens")

# Plot for each year
yrs <- 2018:2023
for (yr in yrs) {
  # Filter by year
  subset_data <- zcta_no_NAs[zcta_no_NAs$year == yr, ]

  # Divide into consistent bins based on the global quantiles
  bins <- cut(subset_data$S1903_C03_001E, breaks = quantiles, labels = FALSE, include.lowest = TRUE)

@@

  # Add legend (use the global quantiles)
  legend("right", inset = c(-.75, 0), legend = quantiles, fill = palette, title = "Median Income ($)", bty = "n", cex = 0.8, xpd = TRUE)
}

# Outer title for the whole set of plots
mtext("Median Household Income by Year and Zip Code in LA", outer = TRUE, cex = 1, font = 2)

@@

## Frequency of Top 10 Crimes

```{r}
crime_frequency <- data.frame(table(crime_la$Crm.Cd.Desc))
top_10_crimes <- crime_frequency[order(-crime_frequency$Freq), ][1:10, ]
```

## Homicide

@@

weekly_counts$CrimesLag <- c(NA, weekly_counts$Crimes[-nrow(weekly_counts)])

@@

summary(ar1_trend_regression)

@@

>>>>

```{r, output=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses

@@

```{r}
# quantiles for the entire dataset (from 2018 to 2023) to divide the median income into groups
qntls <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

# 2x3 plot layout for years 2018-2023, setting margins and outer margins, main title sizes
par(mfrow = c(2, 3), mar = c(1, 1, 1, 5), oma = c(0, 0, 3, 0), cex.main = 1.5)

# color palette for sequential data
palette <- brewer.pal(7, "Greens")

yrs <- 2018:2023

for (yr in yrs) {
  # get just that year's data
  subset_data <- zcta_no_NAs[zcta_no_NAs$year == yr, ]

  # creating bins based on the qtls
  bins <- cut(subset_data$S1903_C03_001E, breaks = qntls, labels = FALSE, include.lowest = TRUE)

@@

  # adding legend
  legend("right", inset = c(-.75, 0), legend = qntls, fill = palette, title = "Median Income ($)", bty = "n", cex = 0.8, xpd = TRUE)
}

# main title for all plots
mtext("Median Household Income by Year and Zip Code in LA", outer = TRUE, cex = 1, font = 2)

@@

## Crime Rates Compared to Median Household Income

```{r}
# using latitude and longitude to obtain the zip codes of the crime locations
# crime_la$year <- as.integer(stringr::str_sub(crime_la$YearWeek, 1, 4))
# id_lat_long_crime_la <- subset(crime_la, select = c("LAT", "LON", "DR_NO", "year"))

# taking a random sample to reduce load on model
#set.seed(123)
#sampled_ids <- id_lat_long_crime_la[sample(1:nrow(id_lat_long_crime_la), 500),]
#rownames(sampled_ids) <- NULL
#zip_codes <- reverse_geocode(sampled_ids, lat = LAT, long = LON, verbose = TRUE)
#write.csv(zip_codes, "sampled_crimes.csv")
zip_codes <- read.csv("sampled_crimes.csv")

zip_codes$ZCTA5CE20 <- stringr::str_extract(zip_codes$address, "\\d{5}")

crime_income_sample <- zip_codes %>% inner_join(unique(subset(zcta_no_NAs, select = c("ZCTA5CE20", "S1903_C03_001E", "year"))), by = c("ZCTA5CE20", "year"))

# this takes a while, so loading in this csv instead

crime_income_sample <- zip_codes %>% inner_join(unique(subset(zcta_no_NAs, select = c("ZCTA5CE20", "S1903_C03_001E", "year"))), by = c("ZCTA5CE20", "year"))

crime_income_per_zip <- crime_income_sample %>% 
  group_by(ZCTA5CE20) %>% 
  summarize(number_of_crimes = n(),
            cali_median_income = mean(S1903_C03_001E)) %>%
  ungroup()

plot(x = crime_income_per_zip$cali_median_income, 
     y = crime_income_per_zip$number_of_crimes, 
     type = "p",
     ylab = "Number of Crimes",  
     xlab = "Median Income ($)",  
     main = "Crime vs. Median Income by ZIP Code",
     bty = "l")
```

### Based on this plot, we do not observe a clear trend between the number of crimes committed and the median income in that zip code. However, this scatterplot only accounts for a small portion of the dataset and averages across pre- and post-pandemic conditions. Future visualizations and analyses could consider pandemic vs non-pandemic years as variables.

# Regression Analyses

@@

```{r}
weekly_counts$CrimesLag <- c(NA, weekly_counts$Crimes[-nrow(weekly_counts)])

@@

summary(ar1_trend_regression)
```
### Interpretation

@@

## Basic Regression Between Median Income and Crimes per Zip Code

```{r}
lm_income_crimes <- lm(number_of_crimes ~ cali_median_income, data = crime_income_per_zip)

summary(lm_income_crimes)
```
### Interpretation

Based on the p-value, we observe that median income is not significant at a 0.05 level, meaning it does not significantly decrease the amount of crimes committed. The r^2 value additionally tells us that the income only explains a small portion of the variation in the number of crimes. As stated with the scatterplot, a larger sample and considering other factors might yield different results.

# Bibliography
